Variability as a better characterization of Shannon entropy

### 2021-01-04 Articles/2019-ShannonEntropy/Submissions/EJP/response.tex

\emph{The use of information in physics, then, does not warrant a fundamental change of perspective in what constitutes a physical object, as some physicists have claimed.}

\noindent which is not at all the same. This simply means, as it is explained, that while matter and energy carry information through their physical processes, they are not themselves ``made'' of information.

\begin{quote}
	even though they wrote in their introduction “we do not discuss the relationship [of the Shannon entropy] to the thermodynamic entropy, which is a separate concept,” in an apparently neutral and unpartisan view of this important issue.
\end{quote}
Thermodynamic entropy is defined in terms of temperature and heat. The paper does not talk about temperature and heat, which are essential if one wants to link thermodynamic entropy to Shannon entropy. That has nothing to do with information being physical.

We added a footnote that reads:

\emph{We will not discuss its relation to thermodynamic entropy. Thermodynamic entropy is defined on the notions of heat and work while the Shannon entropy is defined on the notion of a distribution. The contexts are different and relating them goes beyond the scope of this work.}

@@

\begin{quote}
However, since the underlying implicit thesis to their paper is that the Shannon information is not physical, they should justify this fundamental premise still further.
\end{quote}
That is not our thesis. We simply state, which is not a matter of personal opinion, that the nature of the Shannon entropy depends on the nature of the ``probability'' one is using. In ecology, for example, $p_i$ often represent the fraction of the population of animals that belong to the same species $i$. The Shannon entropy, which is called Shannon diversity index, is an objective (and physically grounded) quantity. In machine learning and Bayesian probability, the $p_i$ represent the degree of belief of an agent that something will happen. The Shannon entropy will represent, by definition, something subjective. We stated this very clearly in the original submission:

\emph{This expression quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on.}

There are multiple comments to that effect throughout the paper. The referee is attributing us an ``implicit'' thesis that is not our own, instead of confronting the explicit ``thesis'' that is repeated throughout the paper.  It is not apparent to us how we could modify the manuscript to avoid that, particularly because he is the only one that has done so.

\begin{quote}
 Their main reference supporting this premise appears to be their ref.[4] – however, this particular reference [4] is itself highly flawed (having both numerous technical and conceptual errors, and also completely ignoring the highly relevant work of E.T. Jaynes – it also a paper not published in an appropriate physics journal – however, I’m not writing here a review of this flawed ref.[4] paper.) But, the fact that the authors of this present paper considered the ref.[4] to be an appropriate reference to support their fundamental assumption about the ‘unphysicality’ of information therefore also unfortunately casts a negative light on their scientific judgement.
\end{quote}
That reference appeared after the sentence ``it is still surrounded by a great deal of confusion''. Therefore it is only used as evidence to show that there is confusion surrounding entropy. This can be taken as the reference showing the confusion, or the reference itself being confused, as the referee argues. In either case, it proves the point: there is confusion.

@@

We find the attacks on the reference completely out of place, inappropriate and distasteful. Ad hominem attacks have no place in the scientific discourse.

\begin{quote}
Hence, their wish to use the word ‘variability’ as an alternative synonym to the word ‘entropy’.
\end{quote}

The referee puts words in our mouth that are not on the page. The use of `variability' for the Shannon formula is justified by this paragraph:

@@

\emph{The first problem is that the formula is attached to different concepts that are specific to each field. [...] In this paper we show that the idea of variability of the elements within a distribution, how diverse the objects are from each other within the collection, can serve as such a general concept. We will therefore use the term \emph{``Shannon variability,''} leaving ``entropy'' to the physical concept.}

So, ``variability'' is used for something that works not only in physics, but also ecology, information theory, machine learning, communication theory, economics... ``Entropy'' is used for the physical concepts in statistical mechanics. That helps the reader understand that some properties of that concept are more general than physics, which is the proper setting to understand the concept. Given that we make a distinction between the two words, ``variability'' and ``entropy'' are \textbf{not} synonyms in our work. That is the whole point. The variability of species in ecology is not a measure of entropy in statistical mechanics.

\begin{quote}
Rather, it is clear to me the Shannon formula is actually the fundamental physical quantity from which all entropy types (be it the information, Gibbs, and Boltzmann entropies etc.) can be derived.
\end{quote}
The referee is expressing a view that, though shared by others, is contested in the literature of the foundations of thermodynamics. For example, see ``Time-reversal and entropy'' by Christian Maes et al, or ``Gibbs and Boltzmann Entropy in Classical and Quantum Mechanics''
Sheldon Goldstein et al. In many modern textbooks of statistical mechanics, the Shannon entropy is not the starting point, rather an approximation that is valid only at equilibrium. See, for example, ``Statistical mechanics in a nutshell'' by Luca Peliti, where the starting point is state counting. We are staying out of that debate as we would find inappropriate, in the context of a paper aimed at scientific education, to push a particular view.

We do believe that part of the confusion surrounding entropy does come from these unresolved contentions. This was clear while doing literature review with some of our students. We therefore added the following sentence in the introduction:

\emph{[... entropy ...] is surrounded by a great deal of confusion. Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium.}

@@

\begin{quote}
The current authors appear to have completely neglected the more recent and highly relevant work written by Jaynes in the 1970s, 1980s, and 1990s, but relied only on his original seminal papers of 1957 and 1965.
\end{quote}

Doing a comparative analysis of this kind is beyond the scope of the paper, which is simply to introduce in a clear way to newcomers what the expression $- \sum p \, \log p$ represents. See, for example, "Compendium of the foundations of classical statistical physics" by Jos Uffink.

\begin{quote}
In particular, the authors here complain that the information entropy is ‘subjective’ in contrast to thermodynamic entropy which is ‘objective’. However, their treatment of this is completely superficial. They need to address Jaynes’ more nuanced and deeper treatment of this issue in his 1992 paper.
\end{quote}
Again, we are in no way disproving what Jaynes says. What the referee (but he is not the only one) is engaging in what philosophers define as ``verbal dispute'', where one disagrees on the meaning of the words rather than on an actual fact of the matter. The problem is what one means by ``subjective''. Jaynes, unfortunately, used language that sometimes seems Bayesian to refer to cases where the experimental setup matters (which is not Bayesian). This language is then taken out of context and used to support claims that do not follow from Jaynes' examples. So our discussion is there to caution the target reader, who is a newcomer to all of this, of these differences.

We have expanded the section to clarify that and also cited the 1992 which does not change anything. [TODO include]

\begin{quote}
	There is an egregious technical error in this paper: In the discussion of the Continuous Distributions (Section IV) after Eq.1 the authors discuss the quantity log(b-a) and suggest that it can be negative for $0<b-a<1$. However, on the assumption that b and a (i.e. the co-ordinate x) in general represent some kind of physical quantity, e.g. length [m]) you simply cannot take the logarithm of a dimensioned quantity! A normalising unit quantity, e.g. $\Delta$, which also defines the granularity or resolution is required to normalise the argument of the logarithm; i.e. we need log\{ (b-a)/$\Delta$ \}. The choice of the granularity (which is a key aspect to any entropic analysis) automatically solves this apparent problem raised by the authors. But they have overlooked even this simple (mathematical and physical) aspect.
\end{quote}
Let $\Delta > b - a > 0$. Then $\frac{b-a}{\Delta} < 1$. Therefore $\log \frac{b-a}{\Delta} < 0$. The resolution does not fix the possible negative value of the Shannon entropy for continuous distributions. The importance of granularity, of unit choice, not only was already present in the introduction, but is literally given in the original submission in paragraph following the one the referee refers to:

@@

\emph{	When considering continuous distributions, we have to remember that the densities are given per unit. [...] A distribution over half a unit interval will correspond to minus one bit, since we would need to ``lose'' one bit of information to widen the variability back to one unit. In other words, the variability is measured by the number of bits needed to identify an element up to one unit.}

Given the complete mis-characterization of the prose and the erroneous claim that somehow the choice of unit/granularity makes the Shannon entropy always positive, we find it impossible to understand whether there is any meaningful item to address. One possible answer is that the referreee does not realize that choosing a unit, in the context of information entropy, is equivalent to choosing the resolution since it changes the zero for entropy... The $\Delta$ \textbf{is} the unit. We added the words ``granularity'' and ``resolution'' in a few places where units are discussed.

@@

The word `perplexing' is used as a rhetorical device, to introduce questions a newcomer might find puzzling, which we answer in the following paragraph. Does the referee really think that means we find the issue perplexing? This would leave us very perplexed...

\begin{quote}
The authors comment that “infinitesimally narrow distributions, like delta functions, do not work well with the Shannon [entropy]”. It is well known that delta functions always require extreme care in their handling, since they are intrinsically unphysical (not conforming to many of the key requirements of physically plausible functions, such as causality and the absence of infinities etc.); so to suggest that the Shannon entropy is somehow inappropriate due to its having apparent problems with the delta function is fatuous – it just means that the delta function is not being appropriately applied. 
\end{quote}
Again, if the referee actually read what we wrote, he would have seen that we say exactly the same thing: that one has to work with functions that have small but still finite support.

\begin{quote}
The authors make a curious comment about the Boltzmann constant, that it “is really just set by the unit system”, as if it isn’t a universal constant and is simply a matter of convention! 
\end{quote}
In the last revision of the SI, universal constants \textbf{are} set to selected values, no longer measured: they are a matter of convention. They are called \textbf{defining constants} because they define the unit system, which is a convention. Before engaging in comments that imply we are incompetent, the referee may want to read the relevant literature.

We added the following footnotes, merely as a pointer for the curious reader. [TODO]

@@

\begin{quote}
There are numerous other contentious and erroneous (careless) statements made by the authors. I’ve merely focussed on the larger, fundamental issues to indicate that this paper is not fit for publication.

Overall, the authors have created a ‘strawman’ argument (by using only partial and weak sources for their argumentation) against the Shannon entropy, so that their paper makes no useful contribution into the literature for a more fundamental understanding of entropy.
\end{quote}

This referee report, while abundant in ad hominen, mischaracterizations and demeaning language, was outstandingly lacking in constructive criticism. This is peer reviewing at its worst. The most likely scenario is that, after a very cursory look, he framed the manuscript in some diatribe the referee has with some other colleagues, stopped reading carefully and engaged in a tirade, assigning to us views that are not our own. This means that most of the comments weren't even directly relevant to our manuscript: even after sifting the report carefully, we found few actionable items.

We urge the referee to conduct himself in a more professional manner. Whether or not variability is a better characterization for the Shannon entropy can be debated rationally without resorting to mockery.

\section*{Referee 3}

\begin{quote}
	I think that the manuscript is well written, well organised and generally speaking, the main topic is interesting also for a not expert reader. 
\end{quote}

We thank the referee for the kind words.

\begin{quote}
The idea of the Shannon entropy as a way to quantify the number of yes/no questions (i.e. bits) that are needed to identify an element within the distribution is not new. However, I appreciate the authors' presentation because this could have validity in the field of Physics Education.
\end{quote}

This is our belief as well and one of the reasons we wrote the article.

\begin{quote}
Although I think that the manuscript deserves to be published, I would like that the authors add some bibliographic references on the topic they deal with but in the field of Physics Education.
\end{quote}

Most of the literature we found in Physics Education focuses on the Gibbs, Boltzmann and Thermodynamic entropies. We added references for those in the first paragraph of the introduction, along with a remark stating that the Shannon entropy itself has received less attention.

\begin{quote}
As a minor issue, I noted a typo at line 27.
\end{quote}

\section*{Referee 4}

\begin{quote}
After reading the paper, Referee’s and Adjudicator’s comments I think that it is proper to send all the correspondence to the Authors. I suggest the Authors to give their replies directly on the pages of the paper: such and similar comments/questions appear in a natural way when one reads the manuscript. The paper (and readers as well) will benefit from modifications along the lines raised by Referees. In fact, I am convinced that  Referee 2 reflections  (the most critical one) will be very useful in this context and can be properly incorporated in the paper. I think that they are well based and clearly explained, however, I do not think that the Authors ‘...have created an ...arguments against the Shannon entropy..’ (as the Referee states). Rather they aim to explain for a wide audience that ‘the Shannon entropy measures the variability of the elements within a given distribution, giving it a crisp intuitive meaning that is general and applicable to all branches of science’ (as the Authors say in their Conclusions).
\end{quote}

We thank the referee for appreciating the goal and the intended audience of the paper. We tried addressing as best we could the points raised by the second referee.

### 2021-01-04 Articles/2019-ShannonEntropy/Submissions/EJP/EJP-ShannonEntropyIsVariability.tex
Preparing response for EJP

The goal of this work is to give a comprehensive introduction to the concept of Shannon entropy, the expression $-\sum p_i \log p_i$, which is well established in such disparate fields as computer science, communication theory, physics\cite{Jaynes1957-1, Grandy} ecology and economics. The general concept of entropy in physics, despite being treated extensively in the physics education literature (e.g. [cite physed]), is still surrounded by a great deal of confusion [cite evidence of confusion]\cite{Swendsen, KishFerry}.Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium. [cite works of out of equilibrium research]  We will focus on the notion of Shannon entropy, which has received less attention in physics education. We will see that the Shannon entropy can be given a crisp and precise characterization and will explore its relation to the Gibbs and Boltzmann entropies in statistical mechanics.\footnote{We will not discuss its relation to thermodynamic entropy: this requires the notions of heat and temperature, which will not be covered in this work.}

@@

When considering continuous distributions, we have to remember that the densities are given per unit. That is, $\rho(x)$ has dimensions $[\frac{1}{x}]$. The variability will be also given relative to the unit. For example, a uniform distribution over a unit interval will correspond to zero variability. A uniform distribution over a two unit interval will correspond to one bit, since we will need one bit to narrow the variability back to one unit. A distribution over half a unit interval will correspond to minus one bit, since we would need to ``lose'' one bit of information to widen the variability back to one unit. In other words, the variability is measured by the number of bits needed to identify an element up to one unit.

This means that infinitesimally narrow distributions, like delta functions, do not work well with the Shannon variability as they would give minus infinite entropy. TODO If one wants entropy over continuous variables to be bounded, then one must work with continuous functions which have a very small but still finite support (i.e.~region where the function is non-zero), which is what we will assume.

@@

The Boltzmann constant $k_B$ should not distract us: its role is simply to allow us to measure temperature in an appropriate unit. If one defines $T = \frac{\partial U}{\partial S}$, and measures temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. Therefore the entropy cannot be expressed as a pure number. However, this is not the only possible definition. Instead of using $T$ as a primary thermodynamic variable, we can use $\beta = \frac{1}{k_B T}$.\footnote{There are a few reasons why one can argue this is a better choice, though ultimately it is still a matter of choice.} In this case, one would define $\beta = \frac{\partial S}{\partial U}$, measure energy in Joules, $\beta$ in inverse Joules and entropy would be dimensionless. As one can do all the thermodynamic calculations using just $\beta$, the constant $k_B$ is really just set by the unit system.

@@

We note that some authors choose to interpret the probability distribution not as coming from repeated independent trials, but as the knowledge one has about the system. This would make the entropy a subjective notion\footnote{Jaynes\cite{Jaynes} called it the anthropomorphic nature of entropy.}, but the associated fluctuations we experimentally observe are indeed objective. We believe the confusion comes from the, correct, realization that the same system, under different conditions, will be described with a different set of thermodynamic variables.\footnote{Jaynes\cite{Jaynes} points out:
\begin{quote}
Consider, for example, a crystal of Rochelle salt. For one set of experiments on it, we work with temperature, pressure, and volume. The entropy can be expressed as some function $S_e(T,P)$. For another set of experiments on the same crystal, we work with temperature, the component $e_{xy}$ of the strain tensor, and the component $P_z$ of electric polarization; the entropy as found in these experiments is a function $S_e(T,e_{xy},P_z)$. It is clearly  meaningless to ask, ``What is the entropy of the crystal?''  unless we first specify the set of parameters which define its thermodynamic state.
\end{quote}}
The choice of the system, in statistical mechanics and thermodynamics, is enough to determine the state space for the microstates, but not enough to determine the set of macrostates that correspond to equilibria. We have to specify the process and the constraints that that process puts on the system. Under a different choice of process and constraints the microstates will fluctuate in different ways since we have changed the dynamics of the system. This is what we stated at the beginning: a distribution, and therefore its variability, is always contingent upon some arbitrary choices. The question, then, is whether the system plus the process plus the set of constraints are enough to determine the entropy, or if we also need to mix in human knowledge. What we find is that they are sufficient, so the entropy does not depend on human knowledge.

>>>>

The goal of this work is to give a comprehensive introduction to the concept of Shannon entropy, the expression $-\sum p_i \log p_i$, which is well established in such disparate fields as computer science, communication theory, physics\cite{Jaynes1957-1, Grandy} ecology and economics. The general concept of entropy in physics, despite being treated extensively in the physics education literature (e.g. [TODO cite physed]), is still surrounded by a great deal of confusion [TODO cite evidence of confusion]\cite{Swendsen, KishFerry}. Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium. [TODO cite works of out of equilibrium research] Partly because the same word ``entropy'' is used to refer to concepts (e.g. topological entropy, thermodynamic entropy, graph entropy, R'{e}nyi entropy, information entropy) that have different definition and various degree of overlap.  We will focus on the notion of Shannon entropy, by which we mean the use of $- \sum_i p_i \log p_i$ in the broadest characterization, which has received less attention in physics education. We will see that the Shannon entropy can be given a crisp and precise characterization and will explore its relation to the Gibbs and Boltzmann entropies in statistical mechanics.\footnote{We will not discuss its relation to thermodynamic entropy. Thermodynamic entropy is defined on the notions of heat and work while the Shannon entropy is defined on the notion of a distribution. The contexts are different and relating them goes beyond the scope of this work. }

@@

When considering continuous distributions, we have to remember that the densities are given per unit. That is, $\rho(x)$ has dimensions $[\frac{1}{x}]$. The variability will be also given relative to the unit. For example, a uniform distribution over a unit interval will correspond to zero variability. A uniform distribution over a two unit interval will correspond to one bit, since we will need one bit to narrow the variability back to one unit. A distribution over half a unit interval will correspond to minus one bit, since we would need to ``lose'' one bit of information to widen the variability back to one unit. In other words, the variability is measured by the number of bits needed to identify an element up to one unit, up to the resolution (i.e. the granularity) given by that unit.

This means that infinitesimally narrow distributions, like delta functions, do not work well with the Shannon variability as they would give minus infinite entropy. If one wants entropy over continuous variables to be bounded, then one must work with continuous functions which  maybe have very small but still finite support (i.e.~region where the function is non-zero), which is what we will assume.

@@

The Boltzmann constant $k_B$ should not distract us: its role is simply to allow us to measure temperature in an appropriate unit.\footnote{In the latest revision of the International Metric System of Units SI,\cite{chyla2011evolution} the Boltzmann constant is one of the defining constants. This means its value is assigned, rather then measured, in a way that properly defines Kelvin, the unit of temperature. The universal constants are essentially used as fixed starting points to define our measurement scales.
}
If one defines $T = \frac{\partial U}{\partial S}$, and measures temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. Therefore the entropy cannot be expressed as a pure number. However, this is not the only possible definition. Instead of using $T$ as a primary thermodynamic variable, we can use $\beta = \frac{1}{k_B T}$. In this case, one would define $\beta = \frac{\partial S}{\partial U}$, measure energy in Joules, $\beta$ in inverse Joules and entropy would be dimensionless. As one can do all the thermodynamic calculations using just $\beta$, the constant $k_B$ is really just set by the unit system.\footnote{As \cite{chyla2011evolution} notes:
\begin{quote}
	Temperature characterizes the average thermal energy of particles in a certain ensemble of particles in the state of thermodynamic equilibrium. Therefore, in principle, one could express this quantity in terms of the unit of energy; this is actually practiced in many fields of physics, especially in statistical physics, where it is quite common to use $\Theta = k_B T$ or $\beta = 1 / k_B T$  as a measure of thermal energy, instead of temperature $T$ measured in Kelvins.
\end{quote}
}

@@

We note that some authors choose to interpret the probability distribution not as coming from repeated independent trials, but as the knowledge one has about the system. This would make the entropy a subjective notion: each observer would have a difference credence distribution, regardless of whether it fits the data, and therefore a different Shannon entropy. A physical quantity, however, must be the same for everybody and the associated fluctuations we experimentally observe are indeed objective. So what is going on?

We believe the confusion comes from the, correct, realization that the same system, under different conditions, will be described with a different set of thermodynamic variables. \footnote{Jaynes\cite{Jaynes} points out:
\begin{quote}
 		Consider, for example, a crystal of Rochelle salt. For one set of experiments on it, we work with temperature, pressure, and volume. The entropy can be expressed as some function $S_e(T,P)$. For another set of experiments on the same crystal, we work with temperature, the component $e_{xy}$ of the strain tensor, and the component $P_z$ of electric polarization; the entropy as found in these experiments is a function $S_e(T,e_{xy},P_z)$. It is clearly  meaningless to ask, ``What is the entropy of the crystal?''  unless we first specify the set of parameters which define its thermodynamic state.
\end{quote}}
The choice of the system, in statistical mechanics and thermodynamics, is enough to determine the state space for the microstates, but not enough to determine the set of macrostates that correspond to equilibria. We have to specify the process and the constraints that that process puts on the system. Under a different choice of process and constraints the microstates will fluctuate in different ways since we have changed the dynamics of the system. This is what we stated at the beginning: a distribution, and therefore its variability, is always contingent upon some arbitrary choices. Since the Gibbs entropy is the variability of a microstate within the distribution identified by the macrostate, it is not a property of the single microstate, it is not a property of the system, but it is a property of the system within that specific process, of the macrostate. Some authors refer to this fact by saying that entropy is not objective \footnote{Jaynes\cite{Jaynes,Jaynes2} called it the anthropomorphic nature of entropy.}.

The issue is that subjective, in the context of probability theory, refers to Bayesian probability, which really means subjective: what ones believes to be true regardless of what experimental evidence there is. This is not at all the same concept as the one outlined before.\footnote{Note how Jaynes\cite{Jaynes2} always put ``subjective'' in double quotes.} The system plus the process (which identifies the set of constraints) determines the ensemble and therefore the entropy, regardless of whether an agent knows what process was used. Even if one wants to give a Bayesian account, then, one has to give it in terms of an agent that has full and exact knowledge of the system and the process. But this is hardly subjective. In our view, this is a case of unfortunate word choice. It is much better to simply state that the entropy is a property of the ensemble, of the distribution, and not of the system. And if understand that the Gibbs entropy is the variability of the microstate within the ensemble, then it is clear that it cannot be a property of the microstate itself, but it is a property of the ensemble.
